Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-12-01 manuscript.tex
Updates from Overleaf

%%%%%%%% Intro comments %%%%%%%%%%%%%%%%%

%Progress in DNA science and engineering has been bolstered by a well-established understanding of sequence-dependent thermodynamics. Thousands of DNA and DNA-complex crystal structures have been captured, and the equilibrium structure of every tetranucleotide motif were simulated in a distributed computational effort \citep{Manuscript2014The2012}. Hybridization thermodynamics can predicted with high fidelity by nearest-neighbor stacking, and these calculations have been experimentally verified to account for mismatched pairs, dangling ends, and other non-native motifs \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}.

% Secondary DNA structures such as hairpins and quadruplexes have also been studied in depth and leverage for nanotechnology applications.

%Furthermore several computational models have been paramaterized to capture these thermodynamic characteristics.

%These studies have fueled increasingly complex structural DNA technologies, but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves \citep{Deluca2020DynamicDevices}. Certain dynamical models for hybridization -- such the nucleation-zipper mechanism and fraying-peeling mechanism for dissociation -- have been well-established for decades (\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition}). Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.

@@

% introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=80mm,
        scale=0.5]{Figs/figs_imp/all_permute_vamps_nolegend.png}
        \caption{A comparison of VAMP-2 scores via 5-fold cross-validation for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordinates. Reciprocal coordinates performed better across the board, but there was no substantial difference between the three featurization methods. This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model.}
        \label{fig:allseq_features_vamp2}
	\end{center}
\end{figure}

@@

% need to discuss tmelt shift
We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 2 nm cutoff. We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation. We noted an average temperature shift of about 4C between experiment and simulation exponentials. While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects. Furthermore, we see an acceleration of about one order of magnitude compared to experiment, although this factor is again sensitive to the exact definition of melting temperature. After accounting for the melting temperature shift and acceleration factor, we saw agreement between the experimental data and simulated relaxation fits. GC-core showed the largest deviation experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.

@@

\begin{figure}  %[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_responses.png}
        \caption{Fitting temperature-dependent trends to the "slow" dissociation mode and "fast" fraying mode detected in experiments. The effective simulation temperature is shifted by 4 degrees celcius to account for systematic differences in simulation and experimental melting temperature. Different scaling factors are applied to fast and response to account for variable coarse-grained acceleration along different degrees of freedom.}
        \label{fig:relaxation-comparison}
	\end{center}
\end{figure}

@@

The fast response -- which Sanstead et. al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables. Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}. All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations \citep{PinamontiTheModels, Zgarbova2014BaseRNA}. Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics. Instead, we measured the fast fraying response by counting frames until either duplex terminal end splits beyond a cutoff. This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal. This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events. Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.

@@

% change state names and scientific notation
% include D here? can see what it would look like...
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=160mm, 
        scale=1]{Figs/figs_imp/allseq_states.png}
        \caption{Nearest neighbor representations, molecular renderings, and sequence-dependent probabilities for PCCA+ macrostates. Shifted state abbreviations indicate the direction of the shifted overhang (5' vs. 3') and the number of shifted motifs relative the native state (1 vs. 2)The dissociated state is not shown but represents the remainder of the stationary probability for each sequence.}
        \label{fig:allseq_table}
	\end{center}
\end{figure}

% potentially get rid of State representation, ask Brennan if they are meaningful?
\begin{figure}[hb!]
	\begin{center}
        \includegraphics[width=160mm, 
        scale=1]{Figs/figs_imp/allseq_transitions.png}
        \caption{Free energies maps and PCCA+ states clusterings are shown in the tICA space in order to enhance visibility in dimensions. Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each state. Arrows indicate the probability of transitioning out of the state within the lag time. Circle areas are proportional to state stability.}
        \label{fig:allseq_transition}
	\end{center}
\end{figure}

Following the analysis pipeline described above, we generate SRV-MSMs for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities. By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S1, 5S2, 3S1, 3S2) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models. Arrows between states show the conditional transition probability to another state within the MSM lag time. We now proceed to analyse the sequence-dependent thermodynamics and kinetic features of each MSM.

@@

Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends\citep{Santalucia2004TM}. The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models \citep{Michele2014EHybridization}. It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN calculations (figure \ref{fig:NN_table}). For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2). We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to macrostate results, however, the 3S1 state is more stable than 5S1 states. Moreover, we do not observe a stable 3S2 cluster for GC-end. This indicates that 5' vs. 3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.

@@

\begin{figure}[ht!] % right place for NN table? could go to SI
	\begin{center} 
        \includegraphics[width=100mm, scale=1]{Figs/figs_imp/shifted_FE_table.png}
        \caption{Comparisons between free energies based on simulation macrostate populations and nearest neighbor calculations. All free energies are normalized such that that the native hybridized state is set to zero. Calculations included dangling end contributions but do not take inert tails effects.}

@@

In addition to state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states. For AT-all, we observe approximately equal probability of transitioning from D to any other state. Once a transition has been made, however, the 5' vs. 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate. Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S1 $\rightarrow$ D) than to more aligned states (5S1 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than the 3S2 $\rightarrow$ 3S1 path to native hybridization. Indeed, for GC-end we observe no significant flux between the 5S2 state and the structurally similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H. The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D. This indicates that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S1 $\rightarrow$ H and 3S1 $\rightarrow$ H once shifted states have formed.

@@

Taken together, our results indicate that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways. A majority of hybridization events occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways pass through some out-of-register state. While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through these states. These observations show that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}. Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and response similarity to other processes such as fraying. To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis \cite{Wyer2014KineticsAT-tracts}. Recent all-atom results identified similar out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers \citep{Xiao2019}. Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand. It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to out-of-register transitions compared to AT motifs. This is possible given that stronger hydrogen binding in GC motifs may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.

@@

In addition to identifying relaxation timescales, the T-jump IR response profile can inform the distribution of dynamics contributing to the response. In particular, the stretched fast response for AT-all and GC-end indicate that there may be a combination of processes occurring on the fast timescale. Furthermore, the GC-end response contains both G:C and A:T vibrational contributions, suggesting that native terminal base fraying may not be the unique source of the response. On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T vibration that points to A:T fraying as the predominant dynamic source. Given that fraying and shifting responses are not clearly distinguishable, we cannot determine whether AT-all and GC-end stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two. In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network (AT-all > GC-end > GC-core ~ GC-mix) and a more stretched fast response.

%The origin of the AT-all and GC-end fast response are most difficult to assign because both shifting and fraying could occur, but the more stretched profile of the fast response for AT-all and GC-end suggests that there may be a combination of processes occurring on the fast timescale. Furthermore, internal displacement mechanisms facilitating direct out-of-register transitions may produced similar responses contributing to the ensemble. Given that it is experimentally challenging to distinguish between these contributions, we cannot confirm whether these responses originate from some population of out-of-register configurations prior the temperature jump, shifting mechanisms during the dissociation process, or some combination of the two. In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network and a more stretched response.

%In examining spectroscopic T-jump signatures for these four sequences, we observed a significantly stretched fast response for AT-all, and a more subtle GC-end stretch relative to GC-core and GC-mix fast responses. For each sequence, the fast response is attributed to terminal base fraying, but stretching indicates a broader distribution of dynamics at this timescale. We expect that fraying would still occur in out-of-register conformations but at a different rate than intact fraying. Furthermore, internal displacement mechanisms facilitating direct out-of-register transitions may produced similar responses contributing to the ensemble. Given that it is experimentally challenging to distinguish between these contributions, we cannot confirm whether these responses originate from some population of out-of-register configurations prior the temperature jump, shifting mechanisms during the dissociation process, or some combination of the two. In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network and a more stretched response. We interrogate these differences by examining high temperature simulations later in this manuscript. (Can discuss more details about fig \ref{fig:fast_stretching} here and potentially include this fig in the main text)?

%It is difficult to distinguish a collection of different starting configurations -- such as out-of-register duplexes -- and their associated dynamics prior to the temperature jump from mechanistic pathways after the jump.
%Although reproducing these stretching responses is not within the scope of this analysis, we take these

@@

The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end. Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4). Previous studies suggests that once key native contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}. Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state. This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs. We observe these events to occur with equal probability on either permutable end of the helix. The F4 state is more accessible from an already bound helix, and once oligos are in F4, they are over 20x more likely to return to H than to D. Thus once a D $\leftrightarrow$ F4 transition has occurred, a F4 $\leftrightarrow$ H transition will likely proceed. On the other hand, H $\leftrightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.

@@

\subsection{GC-mix displays more canonical hybridization and dissociation}

@@

\begin{figure} %[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-mix_transitions.PNG}
        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events. Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until fully dissociation occurs. The SRV-MSM state is shown along the x-axis, indicating the point at which a transition has been determined. Fluctuation between states is common during the transition.}
        \label{fig:GC-mix_transitions}
	\end{center}
\end{figure}

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps.PNG}
        \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end. Color maps show inverse distances between out-of-register ends and complementary ends.}
        \label{fig:GC-end_dmaps}
	\end{center}
\end{figure}

% need to gear this more towards the kinetics
\subsubsection{5s1 state distributions inform differences between AT-all and GC-end kinetics}
While examining molecular renderings, we noticed that a significant proportion of GC-end 5s1 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register. We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding. To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S1 state of both sequences. We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation. This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical "top" and "bottom" of the strands. In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI (Figure \ref{fig:GC-end_dmaps_full}). Diffusion maps built from samples of the 3S1 states are also shown in the SI (Figure \ref{fig:GC-end_dmaps_3prime}).

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-core_dmaps.PNG}
        \caption{Plotting the first two non-trivial diffusion map eigenfunctions. Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
        \label{fig:GC-core_dmaps}
	\end{center}
\end{figure}

\subsubsection{The F4 state interrupts GC-core hybridization and dissociation}  % shift to include this in the overall diffusion map section
%\subsubsection{Diffusion maps show ensemble of frayed states}

>>>>

%These studies have fueled increasingly complex structural DNA technologies, but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves \citep{Deluca2020DynamicDevices}.

@@

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=80mm,
        scale=0.5]{Fig1.pdf}
        \caption{A comparison of VAMP-2 scores via 5-fold cross-validation for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordinates. Reciprocal coordinates performed better across the board, but there was no substantial difference between the three featurization methods. This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model.}
        \label{fig:allseq_features_vamp2}
	\end{center}
\end{figure}

@@

% need to discuss tmelt shift
We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 2 nm cutoff. We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation. We noted an average temperature shift of about 4C between experiment and simulation exponentials. While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects. We used to this temperature shift to inform our equilibrium simulations in the next section. We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature. After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits. At lower temperatures, T-jump slow responses likely contain a mix of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis which considers dissociation alone. GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.

@@

\begin{figure}  %[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Fig2.pdf}
        \caption{Fitting temperature-dependent trends to the "slow" dissociation mode and "fast" fraying mode detected in experiments. The effective simulation temperature is shifted by 4 degrees celcius to account for systematic differences in simulation and experimental melting temperature. Different scaling factors are applied to fast and response to account for variable coarse-grained acceleration along different degrees of freedom.}
        \label{fig:relaxation-comparison}
	\end{center}
\end{figure}

@@

The fast response -- which Sanstead et. al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables. Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}. All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations \citep{PinamontiTheModels, Zgarbova2014BaseRNA}. Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics. Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff. This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal. This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events. Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.

@@

% change state names and scientific notation
% include D here? can see what it would look like...
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=160mm, 
        scale=1]{Fig3.pdf}
        \caption{Nearest neighbor representations, molecular renderings, and sequence-dependent probabilities for PCCA+ macrostates. Shifted state abbreviations indicate the direction of the shifted overhang (5' vs. 3') and the number of shifted motifs relative the native state (1 vs. 2).}
        \label{fig:allseq_table}
	\end{center}
\end{figure}

% potentially get rid of State representation, ask Brennan if they are meaningful?
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=120mm, 
        scale=1]{Fig4.pdf}
        \caption{Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each macrostate. Arrows indicate the probability of transitioning out of the state within the lag time. Circle areas are proportional to state stability.}

        %Free energies maps and PCCA+ states clusterings are shown in the tICA space in order to enhance visibility in dimensions. }
        \label{fig:allseq_transition}
	\end{center}
\end{figure}

Following the analysis pipeline described above, we generate SRV-MSMs for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities. By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S1, 3S1, 5S2, 3S2) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models. Arrows between states show the conditional transition probability to another state within the MSM lag time. We now proceed to analyse the sequence-dependent thermodynamics and kinetic features of each MSM.

@@

Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends\citep{Santalucia2004TM}. The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models \citep{Michele2014EHybridization}. It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}). For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2). We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to macrostate results, however, the 3S1 state is more stable than 5S1 states. Moreover, we do not observe a stable 3S2 cluster for GC-end. This indicates that 5' vs. 3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.

@@

\begin{figure}[ht!] % right place for NN table? could go to SI
	\begin{center} 
        \includegraphics[width=120mm, scale=1]{Fig5.pdf}
        \caption{Comparisons between free energies based on simulation macrostate populations and nearest neighbor calculations. All free energies are normalized such that that the native hybridized state is set to zero. Calculations included dangling end contributions but do not take inert tails effects.}

@@

In addition to state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states. For AT-all, we observe approximately equal probability of transitioning from D to any other state. Once a transition has been made, however, the 5' vs. 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate. Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S1 $\rightarrow$ D) than to more aligned states (5S1 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than the 3S2 $\rightarrow$ 3S1 path towards native hybridization.

For GC-end we observe no significant flux between the 5S2 state and the structurally similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H. The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all. This indicates that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S1 $\rightarrow$ H and 3S1 $\rightarrow$ H once shifted states have formed.

@@

Taken together, our results indicate that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways. A majority of hybridization events occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways and 37\% of H $\rightarrow$ D pathways pass through some out-of-register state. The more shifted 3S2 and 5s2 states are more prone to dissociation and thus contribute to about 5\% of these pathways. While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through any out-of-register states. These observations show that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}. Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and response similarity to other processes such as fraying. To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis \cite{Wyer2014KineticsAT-tracts}. Recent all-atom results identified similar out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers \citep{Xiao2019}. Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand. It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to out-of-register transitions compared to AT motifs. This is possible given that stronger hydrogen binding in GC motifs may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.

@@

In addition to identifying relaxation timescales, the T-jump IR response profile can inform the distribution of dynamics contributing to the response. In particular, the stretched fast response for AT-all and GC-end indicate that there may be a combination of processes occurring on the fast timescale. Furthermore, the GC-end response contains both G:C and A:T vibrational contributions, suggesting that native terminal base fraying may not be the unique source of the response. On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T vibration that points to A:T fraying as the predominant dynamic source. Given that fraying and shifting responses are not clearly distinguishable, we cannot determine whether AT-all and GC-end stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two. Although reproducing stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all > GC-end > GC-core ~ GC-mix) and increased stretching in the fast response. (Can briefly mention further experimental work to isolate shifting responses here?)

% We expect that fraying would still occur in out-of-register conformations but at a different rate than intact fraying.

%It is difficult to distinguish a collection of different starting configurations -- such as out-of-register duplexes -- and their associated dynamics prior to the temperature jump from mechanistic pathways after the jump.

@@

The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end. Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4). Previous studies suggests that once key native contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}. Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state. This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs. We observe these events to occur with equal probability on either permutable end of the helix.

The F4 state is more accessible from an already bound helix, and once oligos are in F4, they are over 20x more likely to return to H than to D. Thus once a D $\leftrightarrow$ F4 transition has occurred, a F4 $\leftrightarrow$ H transition will likely proceed. On the other hand, H $\leftrightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation. We previously noted that frayed dynamics are substantially accelerated in the 3spn2 model, which may lead to the F4 state being sampled more frequently then in experiment. Because GC-core stability is most sensitive to fraying, this effect could cause GC-core to dissociate more frequently and lead to the deviation in slow response we note in figure 2.

@@

\subsection{Heterogeneous base pair distribution promotes canonical hybridization and dissociation}

@@

\begin{figure} %[ht!]
	\begin{center}
        \includegraphics[width=450, scale=1]{Fig6.pdf}
        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events. Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until fully dissociation occurs. The SRV-MSM state is shown along the x-axis, indicating the point at which a transition has been determined. Fluctuation between states is common during the transition.}
        \label{fig:GC-mix_transitions}
	\end{center}
\end{figure}

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=450, scale=1]{Fig7.pdf}
        \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end. Color maps show inverse distances between out-of-register ends and complementary ends.}
        \label{fig:GC-end_dmaps}
	\end{center}
\end{figure}

% need to gear this more towards the kinetics
\subsection{5s1 state distributions reveal differences between AT-all and GC-end kinetics}

While examining molecular renderings, we noticed that a significant proportion of GC-end 5s1 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register. We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding. To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S1 state of both sequences. We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation. This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical "top" and "bottom" of the strands. In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI (Figure \ref{fig:GC-end_dmaps_full}). Diffusion maps built from samples of the 3S1 states are also shown in the SI (Figure \ref{fig:GC-end_dmaps_3prime}).

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=450, scale=1]{Fig8.pdf}
        \caption{Diffusion map representation configurations sampled from the F4 state. The x and y axes represent the the first two non-trivial diffusion map eigenfunctions. Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
        \label{fig:GC-core_dmaps}
	\end{center}
\end{figure}

\subsection{Structural diversity in the F4 state informs fraying dynamics}  % shift to include this in the overall diffusion map section
%\subsubsection{Diffusion maps show ensemble of frayed states}

(might just want to move this to the SI) We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate. We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}). This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs. Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes. We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate. This distance also correlated closely with the second diffusion map mode. Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in "overlap" distance between adjacent A or T bases and the GC core. We defined overlap as the distance between a G/C and the A/T adjacent to its native complement. In these conformations, one of the strands maintains some helical character while the other twists out of place, resulting in WC bonds being obstructed by the oligo backbone. These states represent another potential way in which the hybridization process (or helix reformation) can be kinetically frustrated. We observed this mode to be mostly symmetric, however there is slight tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart. This might be another consequence of differential excluded volume radii in the 3spn2 force field.
